A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-02-23 manuscript.tex
:construction: feature reduction str

Thus, while the final selection of features should always be driven by the research questions and field-specific conventions, for our illustration we chose six features that relate to common psychological research questions and recent works within the field: (1) central tendency, (2) variability, (3) instability, (4) self-similarity, (5) linear trend, and (6) nonlinearity. An overview of the selected features, their substantive interpretations, and mathematical operationalizations is also available in \tblref{tab:esmFeatures}. We provide additional background on the theoretical relevance of the selected features in \appref{app:FeatureBackground}. For each of the six time series features we chose, we chose a mathematical representation that was appropriate for our type of data. We provide a short introduction of each feature below.

@@

Feature selections seek to create a subset of the most important features. The many available approaches differ in how they seek to determine the importance of the individual features \citep{alelyani2014}. Generally speaking, selection methods can be categorized as `filter', `wrapper', or `hybrid' methods \citep{hoque2014}. The filter methods, broadly speaking determine important features by identifying irrelevant features (e.g., because features do not capture much information), and identifying redundant features \citep[e.g., because features capture the same information; e.g., see][]{yu2004}. Wrapper methods on the other hand avoid the feature-based evaluation and focus on the performance of the later model to identify the most important features \citep{kohavi1997}. A wrapper, thus, compares the performance of models with different feature combinations \citep[e.g., see][]{tang2014}. Traditional examples of wrappers are forward selection or backward elimination procedures. Because filter methods might not always perform well and wrapper methods are computationally intensive\footnote{Computationally \textit{k} features could be considered in \(2^k – 1\) possible combinations --- for the example of 80 features that would allow for \(1.20 \times 10^{24}\) (over one septillion) combinations.}, hybrid methods seek to combine the two methods and find a balance between computational effort (i.e., efficiency) and performance \citep[i.e., effectiveness; e.g.,][]{alelyani2014}. Methods might, for example, use a filter step to reduce the size of features considered in a later wrapper step \citep{hsu2011}.

The second general approach to the dimensionality curse has been feature projection. Generally speaking, projection methods seek to transform the many features in such a way that a much smaller number of new variables can accurately capture the variance and structure of the original features \citep[i.e., the data is projected to a lower dimensional space; e.g.,][]{carreira-perpinan1997}. Projection methods are relatively common in psychological research --- including, factor- and principal component analyses \citep{vandermaaten2009}. Commonly, this is achieved using linear transformations \citep[e.g., principal component analysis; see][]{cunningham2015}, or more complex nonlinear transformations \citep[e.g., t-SNE; see]{lee2007}. Generally speaking, feature selection procedures have the benefit that they retain the interpretable feature labels directly and immediately indicate which features were most informative in the sample. Feature projection methods, on the other hand, tend to be more generalized and are more readily available.

@@

\paragraph{Nonlinearity} Changes in psychology are not always linear, instead nonlinearity is is a common feature of psychological time series \citep{hayes2007}. As an example, episodic disorders, such as depression, are most likely best described as non-linear systems \citep{hosenfeld2015}. Similarly, patients in recovery of depression showed sudden changes in the improvement of depression \citep{helmich2020a}. But also substance abuse \citep{boker1998} or attitude changes rarely develop linearly \citep{vandermaas2003}. For psychological researchers, nonlinear developments are, thus, important in two regards. Firstly, nonlinearity as a deviation from linear trends is an important feature defining many psychological time series and secondly, the shape and structure of the nonlinear trend can be essential in understanding the important changes and transitions of human experiences \citep{chow2019}. Conceptually, researchers might accordingly have research question about the type of the development: ``Is the development of anxiety a nonlinear process?'' as well as the shape and structure of the development: ``Does anxiety change as a sudden or smooth transition?'' Accordingly, researchers can capture the type of development with nonlinearity parameters such as the bicoherence metrics \citep{cuddy2009}. The structure of nonlinear trends is often mathematically more difficult to capture \citep{castro-alvarez2022a}. Common ways to capture the trends include the use of polynomial- and differential equation parameters or more broadly to capture how `wiggly' the trend line is (e.g., estimated degrees of freedom of GAM spline models; similar to the number of spikes \citealp[]{caro-martin2018}).

>>>>

Thus, while the final selection of features should always be driven by the research questions and field-specific conventions, for our illustration we chose six features that relate to common psychological research questions and recent works within the field: (1) central tendency, (2) variability, (3) instability, (4) self-similarity, (5) linear trend, and (6) nonlinearity. An overview of the selected features, their substantive interpretations, and mathematical operationalizations is also available in \tblref{tab:esmFeatures}. We provide additional background on the theoretical relevance of the selected features in \appref{app:FeatureBackground}. For each of the six time series features we chose, we chose a mathematical representation that was appropriate for our type of data. We provide a short introduction of each feature below. Beyond the operationalizations we chose for our case study, we provide an R-function that automatically extracts and prepares a large selection of the time series feature operationalizations presented in \tblref{tab:esmFeatures} in our GitHub repository (see the \texttt{featureExtractor} function).

@@

Feature selections seek to create a subset of the most important features. The many available approaches differ in how they seek to determine the importance of the individual features \citep{alelyani2014}. Generally speaking, selection methods can be categorized as `filter', `wrapper', or `hybrid' methods \citep{hoque2014}. The filter methods, broadly speaking determine important features by identifying irrelevant features (e.g., because features do not capture much information), and identifying redundant features \citep[e.g., because features capture the same information; e.g., see][]{yu2004}. Wrapper methods, on the other hand, avoid the feature-based evaluation and focus on the performance of the later model to identify the most important features \citep{kohavi1997}. A wrapper, thus, compares the performance of models with different feature combinations \citep[e.g., see][]{tang2014}. Traditional examples of wrappers are forward selection or backward elimination procedures. Because filter methods might not always perform well and wrapper methods are computationally intensive\footnote{Computationally \textit{k} features could be considered in \(2^k – 1\) possible combinations --- for the example of 80 features that would allow for \(1.20 \times 10^{24}\) (over one septillion) combinations.}, hybrid methods seek to combine the two methods and find a balance between computational effort (i.e., efficiency) and performance \citep[i.e., effectiveness; e.g.,][]{alelyani2014}. Methods might, for example, use a filter step to reduce the size of features considered in a later wrapper step \citep{hsu2011}. Feature selection procedures have the benefit that they retain the interpretable feature labels directly and immediately indicate which features were most informative in the sample.

The second general approach to the dimensionality curse has been feature projection. Broadly speaking, projection methods seek to transform the many features in such a way that a much smaller number of new variables can accurately capture the variance and structure of the original features \citep[i.e., the data is projected to a lower dimensional space; e.g.,][]{carreira-perpinan1997}. Projection methods are relatively common in psychological research --- including, factor- and principal component analyses \citep{vandermaaten2009}. Commonly, this is achieved using linear transformations \citep[e.g., principal component analysis; see][]{cunningham2015}, or more complex nonlinear transformations \citep[e.g., t-SNE; see]{lee2007}. Feature projection methods have been popular because of their widespread availability as well as their high efficiency.

@@

\paragraph{Nonlinearity} Changes in psychology are not always linear, instead nonlinearity is a common feature of psychological time series \citep{hayes2007}. As an example, episodic disorders, such as depression, are most likely best described as non-linear systems \citep{hosenfeld2015}. Similarly, patients in recovery from depression showed sudden changes in the improvement of depression \citep{helmich2020a}. But also substance abuse \citep{boker1998} or attitude changes rarely develop linearly \citep{vandermaas2003}. For psychological researchers, nonlinear developments are, thus, important in two regards. Firstly, nonlinearity as a deviation from linear trends is an important feature defining many psychological time series and secondly, the shape and structure of the nonlinear trend can be essential in understanding the important changes and transitions of human experiences \citep{chow2019}. Conceptually, researchers might accordingly have research question about the type of the development: ``Is the development of anxiety a nonlinear process?'' as well as the shape and structure of the development: ``Does anxiety change as a sudden or smooth transition?'' Accordingly, researchers can capture the type of development with nonlinearity parameters such as the bicoherence metrics \citep{cuddy2009}. The structure of nonlinear trends is often mathematically more difficult to capture \citep{castro-alvarez2022a}. Common ways to capture the trends include the use of polynomial- and differential equation parameters or more broadly to capture how `wiggly' the trend line is (e.g., estimated degrees of freedom of GAM spline models; similar to the number of spikes \citealp[]{caro-martin2018}).
